Galaxy image statistical generator: how to be confident?

### 2024-09-10 paper/main.tex
experiemnt start

Denoising diffusion stochastic models have gain in popularity due to the fact that they get as good as results as GAN and have certain advantages. Notably they do not need any adversarial networks, the mathematical ground is clear and robust, the training is also theoretically well defined. But they use an iterative schema for the generation that can be computationally intensive and slower compared to GAN, and the implementation of stochastic processes may require some theoretical expertise even if this argument can be mitigated. More discussion about the comparison between the two architectures  can be found in this article \cite{dhariwal2021diffusion}. We do not investigate further the mixing of flow-based and diffusion-based models as \citep[e.g.][]{zhang2021diffusion,gong2021interpreting} as well as the use of diffusion models to improve GAN stability as introduced by \cite{Wang2022}. This can be envisaged in a future study.
%
\section{Experiment}
%
Having briefly outlined different architectures of generative models in the previous section it turns out that one aspect in common between diffusion-based and flow-based models consists in the generation process of a $\bm{x}$ sample which starts by sampling a centred isotropic Gaussian distribution $\bm{z}\sim \mathcal{N}(\bm{z},\bm{0},\bm{1})$ (i.e. a white noise) of dimension $d$ equal to the final sample. But a question rises: if we train two generative models of the same kind (i.e, two flow-based, or two diffusion-based) and if we feed them with the same white noise image, does the generative process of the two models give the same final image? If not we can ask ourselves about the nature of the learned process and if we are faced to a real probability density $p(\bm{x})$ sample, or for instance does the result is a sort of mixture of the original dataset images? An answer has been proposed in \cite{kadkhodaie2024generalization} and point out size of the dataset in the passage from \textit{memorization} to \textit{generation} regime. We propose in the next section to replay their experiment using the same \texttt{U-Net} \textit{denoiser} architecture but trained with small sized ($64\times 64$) normalized gray level (1 channel $[0,1]$) centred galaxy images and in comparison use also the \texttt{Glow} flow-based architecture (Fig.~\ref{fig-Glow-archi}).

>>>>

Denoising diffusion stochastic models have gain in popularity due to the fact that they get as good as results as GAN and have certain advantages. Notably they do not need any adversarial networks, the mathematical ground is clear and robust, the training is also theoretically well defined. But they use an iterative schema for the generation that can be computationally intensive and slower compared to GAN, and the implementation of stochastic processes may require some theoretical expertise even if this argument can be mitigated. More discussion about the comparison between the two architectures  can be found in this article \cite{dhariwal2021diffusion}. We do not investigate further the mixing of flow-based and diffusion-based models as \citep[e.g.][]{zhang2021diffusion,gong2021interpreting} as well as the use of diffusion models to improve GAN stability as introduced by \cite{Wang2022}. This can be envisaged in a future study.
%
\section{Experiment}
%
Having briefly outlined different architectures of generative models in the previous section it turns out that one aspect in common between diffusion-based and flow-based models consists in the generation process of a $\bm{x}$ sample which starts by sampling a centred isotropic Gaussian distribution $\bm{z}\sim \mathcal{N}(\bm{z},\bm{0},\bm{1})$ (i.e. a white noise) of dimension $d$ equal to the final sample. But a question rises: if we train two generative models of the same kind, let say two flow-based or two diffusion-based, and if we feed them with the same white noise image, does the generative process of the two models give the same final image? If not we can ask ourselves about the nature of the learned process and if we are faced to a real probability density $p(\bm{x})$ sample or does the result is a sort of mixture of the original dataset images?

An answer has been given in \cite{kadkhodaie2024generalization} and point out the size of the dataset in the passage from \textit{memorization} to \textit{generation} regimes. They used \texttt{CelebA} \citep{Liu2015}, \texttt{LSUN} bedroom \citep{Yu2015} and  \texttt{CelebA HQ} \citep{KarrasALL18} datasets. All dataset images use  gray level single channel ranging in $[0,1]$.  The \texttt{LSUN}  images have been down sampled to $80\times 80$ as well as the \texttt{CelebA} images which also have been further down sampled to $32\times 32$ resolution, while \texttt{CelebA HQ}  have been down sampled to $40\times 40$ resolution.

We propose to replay their experiment using the same \texttt{U-Net} architecture \citep{2015arXiv150504597R}  as \textit{denoiser} consisting of a contracting part and an expending part that allows for multi-scale architecture with scale correlations. Details are found in \cite{kadkhodaie2024generalization}\footnote{Source code:  \url{https://github.com/LabForComputationalVision/memorization_generalization_in_diffusion_models}}. For comparison purposes, we also use the \texttt{Glow} flow-based model implemented\footnote{Source code: \url{https://github.com/rosinality/glow-pytorch}} in \texttt{PyTorch}  \citep{PyTorch2019} for convenience. The architecture has been presented on Figure \ref{fig-Glow-archi}.

Concerning the training dataset, as we need to train two models with up to $100,000$ images each, we have used the SDSS (Sloan Digital Sky Survey) dataset ($306,006$ galaxies) used in \citep{smith2021}\footnoe{See \url{https://github.com/Smith42/astroddpm}} but first all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels and from the $z,g,r$ channels we produce a single  channel in gray levels in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make_lupton_rgb} function with $Q=8$ and $stretch=0.2$ setting}. Moreover the gray level is coded using 5-bits precision.
{\color{red}The training dataset is available in github}. In such scenario,  according to \cite{kadkhodaie2024generalization} results, the transition between memorization to generalisation regimes is expected using this dataset.

The training
